---
title: "proj notes"
output:
  html_document:
    df_print: paged
---

## Setup

```{r mac_pc_setup}
## Mac or PC? -- set git repo path
mac <- TRUE
if (mac){
  git_path <- "~/git-repos/"
} else{
  git_path <- "C:\\Users\\ip98\\Desktop\\git-repos\\"
}

## Utility function to convert mac path to PC
mac_to_pc <- function(path){
  return(gsub("/", "\\\\", path))
}

repo_path <- "datasciencecoursera/course9-dataproducts/proj/"
if (!mac){
  repo_path <- mac_to_pc(repo_path)
}
```

```{knitr_opts}
knitr:::opts_chunk(fig.width = 10, fig.height = 4)
```

```{r load_libraries}
library(tidyverse)
library(stringr)
```

## Intro

Let's use the ["Canada Mortgage and Housing Corporation, average rents for areas with a population of 10,000 and over"](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3410013301) dataset to explore how annual average rents have changed over time in different geographic areas, for different structure and unit types.

The documentation on this dataset isn't great, but I think we can safely assume that the rent values are per month, averaged over the year for multiple datapoints in each category.

```{r load_data}
data_filename <- "data/housing.csv"
if (!mac){
  data_filename <- mac_to_pc(data_filename)
}
data_fullpath <- paste0(git_path, repo_path, data_filename)

housing <- read_csv(data_fullpath)
```

## Clean data

```{r tidy_data}

## Convert colnames to lowercase and replace spaces with underscores
colnames(housing) <- colnames(housing) %>%
  tolower() %>%
  str_replace_all(c(" " = "_"))

## Rename certain columns and select only those we are interested in
housing_sav <- housing ## save dataframe before manipulating
housing <- housing %>%
  rename(year = ref_date) %>%
  rename(avg_monthly_rent = value) %>%
  select(year, geo, type_of_structure, type_of_unit, avg_monthly_rent, terminated) %>%
  drop_na(avg_monthly_rent) %>% ## drop observations with no avg monthly rent value
  mutate_if(sapply(., is.character), as.factor) %>% ## convert char cols to factors 
  mutate(type_of_unit = fct_relevel(type_of_unit, "Two bedroom units", after = 2)) %>% 
  mutate(type_of_structure
          = fct_recode(type_of_structure,
                       "Apt structs of >5 units" 
                          = "Apartment structures of six units and over",
                       "Apt structs of >2 units"
                          = "Apartment structures of three units and over",
                       "Row & apt structs >2 units"
                          = "Row and apartment structures of three units and over",
                       "Row structs >2 units"
                          = "Row structures of three units and over"))
housing_small <- housing %>%
  filter(geo %in% sample(levels(geo), 5))
save(housing, file = "cdn-rent-prices/data/housing.Rdata")
```

Above I saved the tidy data for use with the shiny app.

## Explore data

Now that our data is in a nice, clean form, let's make some plots to look at what we have (and troubleshoot data tidying if necessary).

```{r facet_line_graph}

## Make plots of year vs value, with different colours for types of units, faceted by geographic location and type of structure

## Set variables
select_geo <- "Hamilton, Ontario"
# select_geo <- "Calgary, Alberta"
facet_by_str <- TRUE
if (facet_by_str){
  facet_var <- "type_of_structure"
  colour_var <- "type_of_unit"
} else {
  facet_var <- "type_of_unit"
  colour_var <- "type_of_structure"
}

plot_housing <- housing %>%
  filter(geo == select_geo)
plot_line <- ggplot(plot_housing, aes(x = year,
                              y = avg_monthly_rent,
                              colour = get(colour_var))) +
  geom_line() + 
  facet_grid(cols = vars(get(facet_var))) + 
  labs(title = paste0("Average monthly rents over time for ", select_geo),
       x = "Year",
       y = "Average monthly rent",
       colour = "Type of unit")

print(plot_line)
```

```{r facet_lm}

## Same as line graph, but a linear model
plot_lm <- ggplot(plot_housing, aes(x = year,
                              y = avg_monthly_rent,
                              colour = get(colour_var))) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm") + 
  facet_grid(cols = vars(get(facet_var))) + 
  labs(title = paste0("Average monthly rents over time for ", select_geo),
       x = "Year",
       y = "Average monthly rent",
       colour = "Type of unit")

print(plot_lm)

```