---
title: "proj notes"
output:
  html_document:
    df_print: paged
---

## Setup

```{r mac_pc_setup}
## Mac or PC? -- set git repo path
mac <- FALSE
if (mac){
  git_path <- "~/git-repos/"
} else{
  git_path <- "C:\\Users\\ip98\\Desktop\\git-repos\\"
}

## Utility function to convert mac path to PC
mac_to_pc <- function(path){
  return(gsub("/", "\\\\", path))
}

repo_path <- "datasciencecoursera/course9-dataproducts/proj/"
if (!mac){
  repo_path <- mac_to_pc(repo_path)
}
```

```{r load_libraries}
library(tidyverse)
```

## Intro

Let's use the ["Canada Mortgage and Housing Corporation, average rents for areas with a population of 10,000 and over"](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=3410013301) dataset to explore how annual average rents have changed over time in different geographic areas, for different structure and unit types.

The documentation on this dataset isn't great, but I think we can safely assume that the rent values are per month, averaged over the year for multiple datapoints in each category.

```{r load_data}
data_filename <- "data/housing.csv"
if (!mac){
  data_filename <- mac_to_pc(data_filename)
}
data_fullpath <- paste0(git_path, repo_path, data_filename)

housing <- read_csv(data_fullpath)
```

## Clean data

```{r tidy_data}

## Convert colnames to lowercase
colnames(housing) <- tolower(colnames(housing))
colnames(housing)

## Remove spaces from some colnames of interest
## use stringr?

```
