---
title: "Petal length and width in different types of irises"
author: "Irena Papst"
date: "2019-08-20"
output: ioslides_presentation
---

```{r setup, include=FALSE}
## Knitr setup
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(plotly)
```

## Central question

Do petal length and width differ significantly in different types of irises?

## Data

We use the `iris` dataset (built into R) to explore this question:

```{r load_data, echo = F}
data(iris)
```

```{r explore_data}
## Measurements:
names(iris)
## Species sampled:
levels(iris$Species)
```

## Plot setup

We use the `plotly` package to easily generate an interactive plot that enables us to explore our central question:

```{r plot_setup}
p <- plot_ly(iris,
             x = ~Petal.Length, y = ~Petal.Width,
             type = 'scatter', mode = 'markers',
             color = ~Species)
```

## Plot

```{r plot_display, echo = F}
p
```

## Conclusions

* There is noticeable separation between the setosa cluster and the other two iris types, in terms of petal length and width.

* There is a small amount of overlap between the versicolor and virginica clusters.

* It is plausible that the petal length and width differ significantly for the three types of irises, but this result has not been shown statistically.
