---
title: "Statistical Inference Course Project"
subtitle: "Part 1: Simulation Exercise"
author: Irena Papst
output:
  pdf_document
fontsize: 12 pt
---

```{r opts_main, echo = FALSE}
## Set default knitr chunk options
knitr::opts_chunk$set(dev = 'tikz',
                      echo = FALSE,
                      fig.height = 3)

```

```{r libraries}
## Load necessary libraries
library(purrr)
library(ggplot2)

```

# Overview

We investigate the Central Limit Theorem (CLT), which essentially states that a distribution of sample means (calculated from different samples of the same population) tends toward a normal distribution as the number of sample means collected tends to infinity. What is remarkable about this theorem is that it holds even when the observations sampled are *not* normally distributed themselves. In fact, we will simulate draws from an *exponential* distribution to illustrate the CLT.

# Simulating draws from an exponential distribution

We sample 40 observations from an exponential distribution with $\lambda = 0.2$ and then take the sample means and variances. We repeat this process 1000 times.

```{r sim}
## Set seed for reproducible results
set.seed(15)

## Set parameters
lambda <- 0.2 
n_draw <- 40
n_sim <- 1000

## Simulate draws and calculate sample means and variances
x <- replicate(n_sim, rexp(n_draw, lambda), simplify = FALSE)
sample_means <- map_dbl(x, mean)
s2 <- map_dbl(x, var)

```

# Sample means vs. theoretical mean

The theoretical mean of this exponential distribution is $\mu = 1/\lambda = 1/0.2 = 5$, so we expect the mean of our simulated sample means (*i.e.* the empirical mean) to be close to this value.

We calculate and display both the theoretical and empirical means.

```{r compare_means}
## Calculate theoretical mean
mu <- 1/lambda

## Calculate mean of sample means
xbar <- mean(sample_means)

## Print results
print(paste0("Theoretical mean: ", mu))
print(paste0("Empirical mean: ", round(xbar,3)))

```

To fully convince ourselves that the empirical mean, $\bar{X}$, is close to the theoretical mean, $\mu$, we can quantify just how close these two values are by calculating the relative difference between the empirical and theoretical means, $(\bar{X}-\mu)/\mu$.

```{r rel_diff_means}
## Calculate relative difference
rel_diff <- (xbar-mu)/mu

## Print results
print(paste0("Relative difference: ", round(rel_diff, 5)))

```

This relative difference indicates that the empirical mean is only `r abs(round(rel_diff, 5))*100`% smaller than the sample mean: a very small difference.

## Sample variances vs. theoretical variance

The variance of this exponential distribution is $\sigma^2 = 1/\lambda = 1/0.2 = 5$, so we expect the variance of our simulated sample means (*i.e.* the empirical variance) to be close to the theoretical variance of $\sigma^2/n = 1/(n\lambda^2)$.

We calculate and display both the theoretical and empirical variances.

```{r compare_vars}
## Calculate theoretical variance
sigma2 <- 1/(n_draw*lambda^2)

## Calculate variance of sample means
s2 <- var(sample_means)

## Print results
print(paste0("Theoretical variance: ", sigma2))
print(paste0("Empirical variance: ", round(s2,3)))

```

The relative difference between the empirical variance, $s^2$, and theoretical variance is given by $(s^2-\sigma^2/n)/\sigma^2/n$.

```{r rel_diff_vars}
## Calculate relative difference
rel_diff <- (s2-sigma2)/sigma2

## Print results
print(paste0("Relative difference: ", round(rel_diff, 5)))

```

This relative difference indicates that the empirical variance is only `r abs(round(rel_diff, 5))*100`% smaller than the sample variance: again, a small difference.

# Distribution of sample means

According to the CLT, the distribution of sample means should be normal (with mean $\mu$ and variance $\sigma^2/n$) as the number of samples, $n$, goes to infinity. Thus the distribution of sample means for our 40 draws should be *approximately* normal with those parameters.

To illustrate this point, we plot this histogram of sample means and overlay a normal distribution with mean $\mu = 5$ and variance $\sigma^2/n = 5^2/40$.

```{r dist}
## Convert sample mean data to a data frame
data <- data.frame(sample_means = sample_means)

## Set histogram parameters
n_bins <- 25
fill <- "grey60"

## Plot histogram of sample means with CLT predicted
## normal distribution overtop
subtitle <- "(compared with the theoretically predicted normal distribution)"
p <- ggplot(data = data,
            mapping = aes(x = sample_means)) +
  geom_histogram(aes(y=..density..),
                 bins = n_bins,
                 fill = fill) + 
  labs(x = "Sample mean, $\\bar{X}$",
       y = "Density",
       title = "Simulated distribution of sample means",
       subtitle = subtitle) +
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sqrt(sigma2)),
                colour = "darkorchid1",
                size = 1.2) +
  annotate("text", x = 6, y = 0.4,
           label = "$N(\\mu, \\sigma^2/n)$",
           colour = "darkorchid1")
  
print(p)

```

We see that the theoretically predicted normal distribution matches quite well with the simulated distribution, and so we conclude that the simulated distribution is approximately normal.

\newpage

# Appendix A: Code

The following code was used to generate all of the results presented in this report.

```{r ref.label=knitr::all_labels(), tidy = TRUE, tidy.opts=list(width.cutoff=55), echo = TRUE, eval = FALSE}
```

\newpage

# Appendix B: Session Info

```{r sys_info}
sessionInfo()  
```